Optimizing a PoS Tag Set for Dependency Parsing

### 2016-05-04 submission.tex
Transition to optimization
More on optimization

\section{Introduction}
\label{sec:intro}
PoS tagging is an important preprocessing step for many NLP tasks,
such as dependency parsing \cite{Niv:Hal:Kub:07,Haj:Cia:Joh:09}, named
entity recognition \cite{Tjo:DeM:03} and sentiment analysis
\cite{Wil:Wie:Hof:09}. Whereas much effort has gone into the
development of PoS-taggers, to the effect that this task is often
considered more or less a solved task, considerably less effort has
been devoted to the empirical evaluation of the PoS tag sets
themselves. Error analysis of PoS-taggers indicate that, whereas
tagger improvement though means of learning algorithm or feature
engineering seems to have reached something of a plateau, linguistic
assessment of the distinctions made by the PoS tag set may be a venue
worth investigating further \cite{Man:11}.  Clearly, the utility of a
PoS tag set is tightly coupled with the the downstream task for which
it is performed. Even so, PoS tag sets are usually employed in a ``one
size fits all'' fashion, regardless of the requirements posed by the
task which makes use of this information.

@@

In this paper, we optimize a Norwegian PoS tag set for the task of
data-driven dependency parsing.  We show that the introduction of
morphological distinctions not present in the original tag set, whilst
compromising tagger accuracy, actually leads to significantly improved
parsing accuracy.  In a set of experiments, various morphological
distinctions are introduced and evaluated both intrinsically, i.e. in
terms of PoS-tagging accuracy, and extrinsically, in terms of parsing
accuracy.  Our results show that the introduction of morphological
distinctions not present in the original tag set, whilst compromising
tagger accuracy, actually leads to significantly improved parsing
accuracy. This optimization allows us to bypass the additional step of
morphological analysis, framing the whole preprocessing problem as a
simple tagging task.

@@

\section{Previous/Related work}
\label{sec:prev}
There has been little previous work dedicated specifically to
extrinsic evaluation of the effects of PoS tag sets on downstream
applications.  There has, however, been some work that evaluates the
effects of PoS tag set granularity on PoS tagging. This includes
investigation of the effects of PoS tag sets on tagging of Swedish
\cite{Meg:01,Meg:02} and English \cite{Mac:05}.

@@

\begin{table} \centering
    %\smaller[0.5]
    \begin{tabular}{@{}ll@{}}
        \toprule
        \textbf{Head} & \textbf{Dependent} \\
        \midrule
        Preposition & Prepositional complement \\
        Finite verb & Complementizer \\
        First conjunct & Subsequent conjuncts \\
        Finite auxiliary & Lexical/main verb \\
        Noun & Determiner \\
        \bottomrule
    \end{tabular}
    \caption{Annotation choices in NDT.}
    \label{ndtannotation}
\end{table}

@@

\paragraph{Tag Set Mapping}
In order to carrying out the tag set modifications, we created a mapping that
maps the relevant existing tags to new, more fine-grained tags including more
relevant morphological features for the applicable tokens. A given tag set
modification involves specifying a tag and one or more features to be
concatenated with said tag for all applicable tokens, i.e., tokens that are
assigned said tag and contains the set of features in its morphological
features.

@@

\section{Tag Set Optimization/Experiments}
\label{sec:optimization}
With more fine-grained linguistically
motivated distinctions, we increase the linguistic information represented in
the tags, which may assist the tagger in disambiguating ambiguous and unknown
words, which in turn may aid the parser in recognizing and generalizing
syntactic patterns. However, the addition of more linguistic information to the
tags and thus a more fine-grained tag set will most likely lead to a drop in
tagger accuracy, due to the increase in complexity. The best tagging does not
necessarily lead to the best parse, and it is therefore interesting to
investigate how the tag set modifications may affect the interplay between
tagging and parsing.

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}lllll@{}}
        \toprule
        \textbf{Tag set} & \textbf{MFT} & \textbf{Accuracy} &
        \textbf{LAS} & \textbf{UAS} \\
        \midrule
        Original & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        Full & 85.12\% & 93.46\% & 87.13\% & 90.32\% \\
        Optimized & 89.20\% & 96.85\% & \textbf{88.87\%} & \textbf{91.78\%} \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the optimized tag set, compared to
        the initial tag sets.}
    \label{finalresults}
\end{table}

\section{Optimized Pipeline}
\label{sec:pipeline}

>>>>

\section{Introduction}
\label{sec:intro}
PoS tagging is an important preprocessing step for many NLP tasks,
such as dependency parsing \cite{Niv:Hal:Kub:07,Haj:Cia:Joh:09}, named
entity recognition \cite{Tjo:DeM:03} and sentiment analysis
\cite{Wil:Wie:Hof:09}. Whereas much effort has gone into the
development of PoS-taggers, to the effect that this task is often
considered more or less a solved task, considerably less effort has
been devoted to the empirical evaluation of the PoS tag sets
themselves. Error analysis of PoS taggers indicate that, whereas tagger
improvement though means of learning algorithm or feature engineering seems to
have reached something of a plateau, linguistic assessment of the distinctions
made by the PoS tag set may be a venue worth investigating further
\cite{Man:11}. Clearly, the utility of a PoS tag set is tightly coupled with
the downstream task for which it is performed. Even so, PoS tag sets are
usually employed in a ``one size fits all'' fashion, regardless of the
requirements posed by the task which makes use of this information.

@@

In this paper, we optimize a Norwegian PoS tag set for the task of
data-driven dependency parsing.  We show that the introduction of
morphological distinctions not present in the original tag set, whilst
compromising tagger accuracy, actually leads to significantly improved
parsing accuracy.  In a set of experiments, various morphological
distinctions are introduced and evaluated both intrinsically, i.e. in
terms of PoS tagging accuracy, and extrinsically, in terms of parsing
accuracy.  Our results show that the introduction of morphological
distinctions not present in the original tag set, whilst compromising
tagger accuracy, actually leads to significantly improved parsing
accuracy. This optimization allows us to bypass the additional step of
morphological analysis, framing the whole preprocessing problem as a
simple tagging task.

@@

\section{Previous/Related work}
\label{sec:prev}
There is a considerable lack of previous efforts on extrinsic evaluation of the
effects of PoS tag sets on downstream applications. We will consider(/noe
annet, considerable...consider) papers that evaluate the effects of PoS tag
set granularity on PoS tagging. This includes investigation of the effects of
PoS tag sets on tagging of Swedish \cite{Meg:01,Meg:02} and English
\cite{Mac:05}.

@@

\begin{table} \centering
    %\smaller[0.5]
    \begin{tabular}{@{}ll@{}}
        \toprule
        \textbf{Head} & \textbf{Dependent} \\
        \midrule
        Preposition & Prepositional complement \\
        Finite verb & Complementizer \\
        First conjunct & Subsequent conjuncts \\
        Finite auxiliary & Lexical/main verb \\
        Noun & Determiner \\
        \bottomrule
    \end{tabular}
    \caption{Syntactic annotation choices in NDT.}
    \label{ndtannotation}
\end{table}

@@

\paragraph{Tag Set Mapping}
In order to carry out the tag set modifications, we created a mapping that
takes as input a set of tags and associated morphological features to be
included in said tags. This mapping maps the relevant existing tags to new,
more fine-grained tags including more relevant morphological features for the
applicable tokens. This is done by concatenating the original tag with said
features for all applicable tokens, i.e., tokens that are assigned said tag and
contains the set of features in its morphological features.

@@

\section{Tag Set Optimization/Experiments}
\label{sec:optimization}
By introducing more fine-grained linguistically motivated distinctions in a tag
set, we increase the linguistic information represented in the tags, which may
assist the parser in recognizing and generalizing syntactic patterns. However,
the addition of more linguistic information to the tags and thus a more
fine-grained tag set will most likely lead to a drop in tagger accuracy/tagging
quality due to the increase in complexity. The best tagging does not
necessarily lead to the best parse, and it is therefore interesting to
investigate how the tag set modifications may affect the interplay between
tagging and parsing. As we want to investigate how we can improve the
linguistic quality of a tag set, we will not introduce tag set modifications
which are not linguistically motivated. We will only consider distinctions we
deem linguistically sensible, even if we expect them to impair the
computational tractability; investigating the interplay between these two
considerations is ultimately our goal.

\subsection{Initial/Baseline Experiments}
In an initial round of experiments, we concatenated the tag of each token with
its set of morphological features in order to map the original tag set to a
new, more fine-grained tag set (hereafter referred to as the \emph{full} tag
set). The result of this was a total of 368 tags, which is clearly very
fine-grained. The two initial tag sets, i.e., the original tag set comprising
19 tags and the full tag set comprising 368 tags, thus represent two extremes
in terms of granularity.  To evaluate these tag sets and investigate how the tag
set granularity would affect the performance of tagging and parsing, we trained
and evaluated TnT and Mate on the training and development data, respectively,
on the two tag sets. In Table \ref{inittagseteval}, we report the results of
these experiments. We see that the tagger accuracy drastically drops when going
from the original to the full tag set. The MFT baseline for the original tag
set is 94.14\%, while it drops by almost 9 percentage points to 85.15\% for the
full tag set. TnT reports an accuracy of 97.47\% on the original tag set, which
is reduced to 93.48\% for the full tag set. These results confirm our
hypothesis that the very high linguistic quality in the full, fine-grained tag
set comes at the expense of drops in tagger performance. However, the
additional linguistic information provided by the full tag set improves the
parser performance. With the original tag set, Mate reports an LAS of 87.01\%
and a UAS of 90.19\%, which increases to 87.15\% and 90.39\%, respectively,
when using the full tag set. As we are looking for linguistically informed
distinctions that improve the syntactic parsing, these results are promising
and serve to indicate that additional morphological information assists the
syntactic parsers, motivating the optimization of the existing PoS tag set.

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}lllll@{}}
        \toprule
        \textbf{Tag set} & \textbf{MFT} & \textbf{Accuracy} &
        \textbf{LAS} & \textbf{UAS} \\
        \midrule
        Original & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        Full & 85.15\% & 93.48\% & \textbf{87.15\%} & \textbf{90.39\%} \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the two initial tag sets.}
    \label{inittagseteval}
\end{table}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}lllll@{}}
        \toprule
        \textbf{Tag set} & \textbf{MFT} & \textbf{Accuracy} &
        \textbf{LAS} & \textbf{UAS} \\
        \midrule
        Original & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        Full & 85.15\% & 93.48\% & 87.15\% & 90.39\% \\
        Optimized & 89.20\% & 96.85\% & \textbf{88.87\%} & \textbf{91.78\%} \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the optimized tag set,
        compared to the initial tag sets.}
    \label{finalresults}
\end{table}

%\section{Optimized Pipeline}
%\label{sec:pipeline}
